---
title: "Prognostic markers of inflammation in endometrioid and clear cell ovarian cancer"
subtitle: "Descriptive analysis"
author: "Barbara Hernando"
output: 
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(kableExtra)

# WORKING DIRECTORY
PATH <- dirname(rstudioapi::getSourceEditorContext()$path)

# LIBRARIES
library(readxl)
library(survival)
library(dplyr)
```

## LOAD DATA

```{r load_data, warning=FALSE}
#Read localized and advanced data 
BDL <- read_excel(paste0(PATH,"/data/BDL.xlsx"))
BDA <- read_excel(paste0(PATH,"/data/BDA.xlsx"))

#Join both datasets
BD <- rbind(BDA, BDL)
```

## PROCESS VARIABLES

```{r process_data}
#SURVIVAL
#Calculate survival in days
diag<-gsub(x=BD$DIAG,pattern=" UTC",replacement="",fixed=T)
dead<-gsub(x=BD$LASTv,pattern=" UTC",replacement="",fixed=T)
BD$Survival <- as.Date(dead, format="%Y-%m-%d")-as.Date(diag, format="%Y-%m-%d") #in days
BD$Survival <- BD$Survival/365 #in years
BD$Survival <- BD$Survival*12 #in months

#RESIDUALS --> 0=No / 1=Yes
BD$RESIDUALsD <- factor(BD$RESIDUALsD, labels=c("No","Yes"))

#SUBTYPE --> 0=EOC / 1=CCOC
BD$SUBTYPE <- factor(BD$SUBTYPE, labels=c("EOC","CCOC"))

#AGE
BD$AGE <- as.numeric(BD$AGEY)

#FIGO --> 0=localized / 1=advanced
BD$FIGO <- factor(BD$FIGOL, labels=c("Stage I/II", "Stage III/IV"))

#ENDOMETROSIS -->0=No / 1=Yes
BD$ENDOMETRIOSIS[BD$ENDOMETRIOSIS==0]<-2 #0 is changed to 2 because the reference category is Yes
BD$ENDOMETRIOSIS <- factor(BD$ENDOMETRIOSIS, labels=c("Yes", "No"))
```

## SPLIT DATA BY HISTOLOGICAL SUBTYPE

```{r subtypes}
CC <- BD[BD$SUBTYPE == "CCOC",] #clear cell == 1
ED <- BD[BD$SUBTYPE == "EOC",] #endometrioid == 0
```

## DESCRIPTION ANALYSES
### 1. Sample size per subtype and stage
```{r dist1}
table <- xtabs(~BD$SUBTYPE+BD$FIGO, data=BD)
ftable(table)
summary(table)
```
### 2. AGE
```{r dist2}
#All population
median(BD$AGE, na.rm=T)
max(BD$AGE, na.rm=T)
min(BD$AGE, na.rm=T)
#EOC
median(ED$AGE, na.rm=T)
max(ED$AGE, na.rm=T)
min(ED$AGE, na.rm=T)
#CCOC
median(CC$AGE, na.rm=T)
max(CC$AGE, na.rm=T)
min(CC$AGE, na.rm=T)
```
### 3. Residual disease per subtype and stage
```{r dist3}
table <- xtabs(~BD$FIGO+BD$SUBTYPE+BD$RESIDUALs, data=BD)
ftable(table)
summary(table)
```

### 4. ENDOMETRIOSIS distribution per subtype and stage
```{r dist4}
table <- xtabs(~BD$FIGO+BD$SUBTYPE+BD$ENDOMETRIOSIS, data=BD)
ftable(table)
summary(table)
```

### 5. CD8TO distribution per subtype
```{r dist5}
BD$CD8TO[BD$CD8TO==9]<-NA
table <- xtabs(~BD$SUBTYPE+BD$CD8TO, data=BD)
ftable(table)
summary(table)
```
### 6. CD8iTO distribution per subtype
```{r dist6}
BD$CD8iTO[BD$CD8iTO==9]<-NA
table <- xtabs(~BD$SUBTYPE+BD$CD8iTO, data=BD)
ftable(table)
summary(table)
```
### 7. CD3TO distribution per subtype
```{r dist7}
BD$CD3TO[BD$CD3TO==9]<-NA
table <- xtabs(~BD$SUBTYPE+BD$CD3TO, data=BD)
ftable(table)
summary(table)
```

### 8. CD3iTO distribution per subtype
```{r dist8}
BD$CD3iTO[BD$CD3iTO==9]<-NA
table <- xtabs(~BD$SUBTYPE+BD$CD3iTO, data=BD)
ftable(table)
summary(table)
```

### 9. NLR distribution per stage in EOC
```{r dist9}
median <- median(as.numeric(ED$NL), na.rm=TRUE)
ED$NLR <- cut(as.numeric(ED$NL), breaks=c(-Inf, median, Inf), labels=c("Low","High"))
table <- xtabs(~ED$FIGO+ED$NLR, data=ED)
ftable(table)
summary(table)
```

### 10. NLR distribution per stage in CCOC
```{r dist10}
median <- median(as.numeric(CC$NL), na.rm=TRUE)
CC$NLR <- cut(as.numeric(CC$NL), breaks=c(-Inf, median, Inf), labels=c("Low","High"))
table <- xtabs(~CC$FIGO+CC$NLR, data=CC)
ftable(table)
summary(table)
```
