---
title: "Prognostic markers of inflammation in endometrioid and clear cell ovarian cancer"
subtitle: "Multivariate survival models"
author: "Barbara Hernando"
output: 
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(kableExtra)

# WORKING DIRECTORY
PATH <- "C:/Users/bhernando/Desktop/CNIO/Projects/MariaProject/Alex"

# LIBRARIES
library(readxl)
library(survival)
library(dplyr)
```

## LOAD DATA

```{r load_data, warning=FALSE}
#Read localized and advanced data 
BDL <- read_excel(paste0(PATH,"/BDL.xlsx"))
BDA <- read_excel(paste0(PATH,"/BDA.xlsx"))

#Join both datasets
BD <- rbind(BDA, BDL)
```

## PROCESS VARIABLES

```{r process_data}
#SURVIVAL
#Calculate survival in days
diag<-gsub(x=BD$DIAG,pattern=" UTC",replacement="",fixed=T)
dead<-gsub(x=BD$LASTv,pattern=" UTC",replacement="",fixed=T)
BD$Survival <- as.Date(dead, format="%Y-%m-%d")-as.Date(diag, format="%Y-%m-%d") #in days
BD$Survival <- BD$Survival/365 #in years
BD$Survival <- BD$Survival*12 #in months

#RESIDUALS --> 0=No / 1=Yes
BD$RESIDUALsD <- factor(BD$RESIDUALsD, labels=c("No","Yes"))

#SUBTYPE --> 0=EOC / 1=CCOC
BD$SUBTYPE <- factor(BD$SUBTYPE, labels=c("EOC","CCOC"))

#AGE
BD$AGE <- as.numeric(BD$AGEY)

#FIGO --> 0=localized / 1=advanced
BD$FIGO <- factor(BD$FIGOL, labels=c("Stage I/II", "Stage III/IV"))

#ENDOMETROSIS -->0=No / 1=Yes
BD$ENDOMETRIOSIS[BD$ENDOMETRIOSIS==0]<-2 #0 is changed to 2 because the reference category is Yes
BD$ENDOMETRIOSIS <- factor(BD$ENDOMETRIOSIS, labels=c("Yes", "No"))

#CD3T OTTA (intratumoral) --> none/low <=2 / medium/high >2
BD$CD3TO <- cut(as.numeric(BD$CD3T), breaks=c(-Inf, 2, Inf), labels=c("None/Low","Medium/High"))

#CD3iT OTTA (intraepithelial) --> none/low <=2 / medium/high >2
BD$CD3iTO <- cut(as.numeric(BD$CD3iT), breaks=c(-Inf, 2, Inf), labels=c("None/Low","Medium/High"))

#CD8T OTTA (intratumoral) --> none/low <=2 / medium/high >2
BD$CD8TO <- cut(as.numeric(BD$CD8T), breaks=c(-Inf, 2, Inf), labels=c("None/Low","Medium/High"))

#CD8iT OTTA (intratumoral) --> none/low <=2 / medium/high >2
BD$CD8iTO <- cut(as.numeric(BD$CD8iT), breaks=c(-Inf, 2, Inf), labels=c("None/Low","Medium/High"))
```

### SPLIT DATA BY HISTOLOGICAL SUBTYPE
The database is split by histological subtype. Moreover, we have removed the advanced tumours in ECC for survival analyses because there are only 13 samples
```{r subtypes}
CC <- BD[BD$SUBTYPE == "CCOC",] #clear cell == 1
ED <- BD[BD$SUBTYPE == "EOC",] #endometrioid == 0
ED <- ED[ED$FIGO != "Stage III/IV",]
```

### MULTIVARIATE SURVIVAL ANALYSES

#### 1. Intratumoral CD8

##### 1.1. Clear cell ovarian carcinomas (CCOC)

```{r km1, fig1, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit <- survfit(Surv(Survival, DEATHo) ~ CD8TO+FIGO, data = CC)
cols4=c("firebrick1", "dodgerblue1", "firebrick4", "dodgerblue4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols4)
print(p)
```

```{r hr1,  fig2, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ CD8TO+FIGO+AGE+RESIDUALsD, data = CC)
p <- ggforest(fit.coxph, data = CC, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

##### 1.2. Endometroid ovarian carcinomas (EOC)

```{r km2, fig3, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit <- survfit(Surv(Survival, DEATHo) ~ CD8TO, data = ED)
cols2=c("firebrick1", "firebrick4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols2)
print(p)
```

```{r hr2,  fig4, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ CD8TO+AGE, data = ED)
p <- ggforest(fit.coxph, data = ED, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

#### 2. Intraepithelial CD8:

##### 2.1. Clear cell ovarian carcinomas (CCOC)
```{r km3, fig5, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit <- survfit(Surv(Survival, DEATHo) ~ CD8iTO+FIGO, data = CC)
cols4=c("firebrick1", "dodgerblue1", "firebrick4", "dodgerblue4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols4)
print(p)
```

```{r hr3,  fig6, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ CD8iTO+FIGO+AGE+RESIDUALsD, data = CC)
p <- ggforest(fit.coxph, data = CC, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

##### 2.2. Endometroid ovarian carcinomas (EOC)
```{r km4, fig7, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit <- survfit(Surv(Survival, DEATHo) ~ CD8iTO, data = ED)
cols2=c("firebrick1", "firebrick4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols2)
print(p)
```

```{r hr4,  fig8, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ CD8iTO+AGE, data = ED)
p <- ggforest(fit.coxph, data = ED, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

#### 3. Intratumoral CD3:

##### 3.1. Clear cell ovarian carcinomas (CCOC)

```{r km5, fig9, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit <- survfit(Surv(Survival, DEATHo) ~ CD3TO+FIGO, data = CC)
cols4=c("firebrick1", "dodgerblue1", "firebrick4", "dodgerblue4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols4)
print(p)
```

```{r hr5,  fig10, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ CD3TO+FIGO+AGE+RESIDUALsD, data = CC)
p <- ggforest(fit.coxph, data = CC, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

##### 3.2. Endometroid ovarian carcinomas (EOC)

```{r km6, fig10, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit <- survfit(Surv(Survival, DEATHo) ~ CD3TO, data = ED)
cols2=c("firebrick1", "firebrick4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols2)
print(p)
```

```{r hr6,  fig11, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ CD3TO+AGE, data = ED)
p <- ggforest(fit.coxph, data = ED, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

#### 4. Intraepithelial CD3:

##### 4.1. Clear cell ovarian carcinomas (CCOC)

```{r km7, fig12, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit <- survfit(Surv(Survival, DEATHo) ~ CD3iTO+FIGO, data = CC)
cols4=c("firebrick1", "dodgerblue1", "firebrick4", "dodgerblue4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols4)
print(p)
```

Then, we performed a cox proportional-hazards model corrected by other confounding variables. We have only corrected by the variables included in the OTTA study
```{r hr7,  fig13, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ CD3iTO+FIGO+AGE+RESIDUALsD, data = CC)
p<- ggforest(fit.coxph, data = CC, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

##### 4.2. Endometroid ovarian carcinomas (EOC)

```{r km8, fig14, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit <- survfit(Surv(Survival, DEATHo) ~ CD3iTO, data = ED)
cols2=c("firebrick1", "firebrick4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols2)
print(p)
```

```{r hr8,  fig15, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ CD3iTO+AGE, data = ED)
p <- ggforest(fit.coxph, data = ED, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

#### 5. NLR:
Threshold used to categorize samples is the median

##### 5.1. Clear cell ovarian carcinomas (CCOC)

```{r km9, fig16, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
median <- median(as.numeric(CC$NL), na.rm=TRUE)
CC$NLR <- cut(as.numeric(CC$NL), breaks=c(-Inf, median, Inf), labels=c("Low","High"))
fit <- survfit(Surv(Survival, DEATHo) ~ NLR+FIGO, data = CC)
cols4=c("firebrick1", "dodgerblue1", "firebrick4", "dodgerblue4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols4)
print(p)
```

```{r hr9,  fig17, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ NLR+FIGO+AGE+RESIDUALsD, data = CC)
p<- ggforest(fit.coxph, data = CC, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

##### 5.2. Endometroid ovarian carcinomas (EOC)

```{r km10, fig18, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
median <- median(as.numeric(ED$NL), na.rm=TRUE)
ED$NLR <- cut(as.numeric(ED$NL), breaks=c(-Inf, median, Inf), labels=c("Low","High"))
fit <- survfit(Surv(Survival, DEATHo) ~ NLR, data = ED)
cols2=c("firebrick1", "firebrick4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols2)
print(p)
```

```{r hr10,  fig19, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ NLR+AGE, data = ED)
p <- ggforest(fit.coxph, data = ED, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

#### 6. ENDOMETRIOSIS:

##### 6.1. Clear cell ovarian carcinomas (CCOC)

```{r km11, fig20, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit <- survfit(Surv(Survival, DEATHo) ~ ENDOMETRIOSIS+FIGO, data = CC)
cols4=c("firebrick1", "dodgerblue1", "firebrick4", "dodgerblue4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols4)
print(p)
```

```{r hr11,  fig21, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ ENDOMETRIOSIS+FIGO+AGE+RESIDUALsD, data = CC)
p<- ggforest(fit.coxph, data = CC, main = "Hazard ratio", fontsize = 0.7)
print(p)
```

##### 6.2. Endometroid ovarian carcinomas (EOC)

```{r km12, fig22, fig.height = 6, fig.width = 14, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit <- survfit(Surv(Survival, DEATHo) ~ ENDOMETRIOSIS, data = ED)
cols2=c("firebrick1", "firebrick4")
p <- ggsurvplot(fit,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(base_size = 8), # Change ggplot2 theme
          #xlim=c(0,180),
          break.time.by=50,
          palette = cols2)
print(p)
```

```{r hr12,  fig23, fig.height = 4, fig.width = 10, fig.align = "center", eval=TRUE, echo=FALSE, warning=FALSE}
fit.coxph <- coxph(Surv(Survival, DEATHo) ~ ENDOMETRIOSIS+AGE, data = ED)
p <- ggforest(fit.coxph, data = ED, main = "Hazard ratio", fontsize = 0.7)
print(p)
```
